---
title: "Input Data Examination"
author: "Rebecca Thomson"
date: "2024-07-18"
output: html_document
---
Libraries used:
```{r Libraries}
library(tidyverse)
#library(Rtools)
library(redist) #Map Analysis
library(dplyr)
library(ggplot2)
library(alarmdata) #State graph/mapping data
library(ggredist) #Alarm data mapping
```

## Load Data from Package and file

The three states we will be analyzing will be:
Illinois
Indiana
Michigan

Please note that some parts of this analysis are very computationally expensive.  When this is the case, each state has been divided into a separate chunk.  It is advisable to only run these chunks as needed.

```{r load}

IL_state_map<-alarm_50state_map('IL')
#MI_state_map<-alarm_50state_map('MI')
#IN_state_map<-alarm_50state_map('IN')

IL_state<-read.csv(file='il_2020_vtd.csv')
#IN_state<-read.csv(file='in_2020_vtd.csv')
#MI_state<-read.csv(file='mi_2020_vtd.csv')
```

## Check dimension, NAs

```{r basic}

print("Checking Illinois:")
dim_to_check<-dim(IL_state_map)
print("The dimensions of the Illinois input data is: ")
print(dim_to_check)
# count total missing values 
print("Count of total missing values in Illinois Data:  ")
sum(is.na(IL_state))

# find location of missing values
print("Position of missing values ")
which(is.na(IL_state))
# 
# print("Checking Michigan:")
# dim_to_check<-dim(MI_state_map)
# print("The dimensions of the Michigan input data is: ")
# print(dim_to_check)
#  
# # count total missing values 
# print("Count of total missing values in Michigan Data:  ")
# sum(is.na(MI_state))
# 
# # find location of missing values
# print("Position of missing values ")
# which(is.na(MI_state))


```


## Input Data Analysis:
Some basic review of important input variables from the data.
```{r data info texas}
#Texas
# par(mfrow=c(1,2))
# hist(TX_state$pop, xlab="Population Histogram by sub-area")
# boxplot(TX_state$pop, xlab="Population Boxplot by sub-area")
# paste("The range of the population per node is from ", min(TX_state$pop), " to ",max(TX_state$pop))
# paste("The mean of the population is ", mean(TX_state$pop))
# par(mfrow=c(1,2))
# texas_ratio_18<-(TX_state$adv_18/(TX_state$adv_18+TX_state$arv_18))
# boxplot(texas_ratio_18, xlab="Ratio of Democrats Boxplot by sub-area")
# texas_ratio_race<-(1-TX_state$pop_white/(TX_state$pop))
# boxplot(texas_ratio_race, xlab="Ratio of Non-white Population Boxplot by sub-area")
# par(mfrow=c(1,2))
# plot(texas_ratio_18,texas_ratio_race, xlab="Ratio of Democratic Voters", ylab="Ratio of non-White Population")
# plot(texas_ratio_18,TX_state_map$pop, xlab="Ratio of Democratic Voters", ylab="Total Population")
```
```{r data info Illinois}
areas= as.numeric(units::set_units(sf::st_area(IL_state_map$geometry),mi^2))
illinois_ratio_18<-(IL_state$adv_18/(IL_state$adv_18+IL_state$arv_18))
illinois_ratio_race<-(1-IL_state$pop_white/(IL_state$pop))
Illinois_pop_density<-IL_state$pop/areas
all_ones<-factor(IL_state$state)
IL_data<-data.frame(areas,illinois_ratio_18,illinois_ratio_race,Illinois_pop_density,all_ones)

par(mfrow=c(1,2))
hist(IL_state$pop, xlab="Population by node", main = "Total Population by node")
boxplot(IL_state$pop, xlab="Population Boxplot by node")
paste("The range of the population per node is from ", min(IL_state$pop), " to ",max(IL_state$pop))
paste("The mean of the population is ", mean(IL_state$pop))
par(mfrow=c(1,2))

boxplot(illinois_ratio_18, xlab="Ratio of Democrats Boxplot by node")

boxplot(illinois_ratio_race, xlab="Ratio of Non-white Population Boxplot by node")
par(mfrow=c(1,2))
plot(illinois_ratio_18,illinois_ratio_race, xlab="Ratio of Democratic Voters", ylab="Ratio of non-White Population", main = "Minority Ratio to Voting History Ratio per node")
abline(lm(illinois_ratio_race~illinois_ratio_18), col='red')
plot(illinois_ratio_18,IL_state_map$pop, xlab="Ratio of Democratic Voters", ylab="Total Population",main = "Voting History Ratio to Total Population per node")
abline(lm(IL_state_map$pop~illinois_ratio_18), col='red')

plot(illinois_ratio_18,Illinois_pop_density, xlab="Ratio of Democratic Voters", ylab="Node Population Density",)
#abline(lm(Illinois_pop_density$pop~illinois_ratio_18), col='red')
```



## Input Data Mapping:

Mapping the import data of Illinois for better visualization
```{r map01-illinois}
#Illinois
areas= as.numeric(units::set_units(sf::st_area(IL_state_map$geometry),mi^2))
plot(IL_state_map, adj=T)

#plot(IL_state_map, adj=T, xlim= c(0,1), ylim= c(0,1))
IL_map_plot<-IL_state_map |>
  ggplot() +
  geom_sf(aes(fill = ndv / (nrv+ndv)))+
  scale_fill_party_c() +
  theme_map()
IL_map_plot


```
## Redist Maps:
This will simulate 2 runs of 1,000 District maps using the redist package.  These random sampling 
 
```{r redist maps-illinois}
#Illinois
illinois_plans<-redist_smc(IL_state_map,4000,compactness = 1,runs = 2 )
print(illinois_plans)

```
```{r redist maps-michigan}
# #Michigan
# michigan_plans<-redist_smc(MI_state_map,500,compactness = 1,runs = 2 )
# print(michigan_plans)

```
```{r redist maps-indiana}
# #Indiana
# indiana_plans<-redist_smc(IN_state_map,500,compactness = 1,runs = 2 )
# print(indiana_plans)

```
## Plot a few examples of the redist plans
These are a few of the random districting plans made by the redist package.

```{r sample plots-illinois}
redist.plot.plans(illinois_plans, draws=c("cd_2020","1","2","300","2000","3001"), shp=IL_state_map)
```

```{r sample plots-indiana}
#redist.plot.plans(indiana_plans, draws=1:6, shp=IN_state_map)
```
```{r sample plots-michigan}
#redist.plot.plans(michigan_plans, draws=1:6, shp=MI_state_map)
```

## Analyze existing districts to  the redist plans


```{r 2020 comparison - illinois}
#Matching district numbers

illinois_plans_2020<-match_numbers(illinois_plans,IL_state_map$cd_2020)

IL_county_perims2020<- prep_perims(IL_state_map, IL_state_map$adj)

#we will be analyze the population deviation, the perimeter-based compactness measure, and the fraction of minority voters and two-party Democratic vote share by district.

illinois_plans_2020 = illinois_plans_2020 %>%
  mutate(pop_dev = abs(total_pop / get_target(IL_state_map)-1),
         comp=comp_polsby(pl(), IL_state_map,perim_df = IL_county_perims2020),
         pct_min = group_frac(IL_state_map, vap-vap_white,vap),
         pct_dem= group_frac(IL_state_map,adv_18,adv_18+arv_18))

print(illinois_plans_2020)

```

```{r 2020 comparison - indiana}
#Matching district numbers
# 
# indiana_plans_2020<-match_numbers(indiana_plans,IN_state_map$cd_2020)
# 
# IN_county_perims2020<- prep_perims(IN_state_map, IN_state_map$adj)
# 
# #we will be analyze the population deviation, the perimeter-based compactness measure, and the fraction of minority voters and two-party Democratic vote share by district.
# 
# indiana_plans_2020 = indiana_plans_2020 %>%
#   mutate(pop_dev = abs(total_pop / get_target(IN_state_map)-1),
#          comp=comp_polsby(pl(), IN_state_map,perim_df = county_perims2020),
#          pct_min = group_frac(IN_state_map, vap-vap_white,vap),
#          pct_dem= group_frac(IN_state_map,adv_18,adv_18+arv_18))
# 
# print(indiana_plans_2020)

```
```{r 2020 comparison - Michigan}
#Matching district numbers
# 
# michigan_plans_2020<-match_numbers(michigan_plans,MI_state_map$cd_2020)
# 
# #Adding summary statistics for both Census
# 
# #MI_county_perims2020<- prep_perims(MI_state_map, MI_state_map$adj)
# 
# #we will be analyze the population deviation, the perimeter-based compactness measure, and the fraction of minority voters and two-party Democratic vote share by district.
# 
# michigan_plans_2020 = michigan_plans_2020 %>%
#   mutate(pop_dev = abs(total_pop / get_target(MI_state_map)-1),
#          comp=comp_polsby(pl(), MI_state_map,perim_df = county_perims2020),
#          pct_min = group_frac(MI_state_map, vap-vap_white,vap),
#          pct_dem= group_frac(MI_state_map,adv_18,adv_18+arv_18))
# 
# print(michigan_plans_2020)

```

## Checking algorithm's Diagnostics
These are the diagnostics recommended by the package.  Our first problem is that the SMC runs did not converge with 500 samples, so the original texas_plans had to be re-done with more samples.

Other factors that must be checked are:
low effective samples
very low acceptance rates (less than 1%)
large std. devs. of the log weights (more than ~3)
low numbers of unique plans.
R-hat values for summary should be between 1 and 1.05

```{r Diagnostics-illinois}
summary(illinois_plans_2020)

```
```{r Diagnostics-indiana}
#summary(indiana_plans_2020)

```
```{r Diagnostics-michigan}
#summary(michigan_plans_2020)

```
## Histograms of random plans
These are a few of the random districting plans made by the redist package.

```{r hist-texas}
#hist(plans_diversity(texas_plans_2020), breaks=24)
```
```{r hist-illinois}
hist(plans_diversity(illinois_plans_2020), breaks=24)
```

```{r post-run analysis-illinois 01}
m_illinois_plans_2020 = illinois_plans_2020 %>%
  mutate(Compactness = comp_polsby(pl(),IL_state_map),
         Population_deviation= plan_parity(IL_state_map),
         Democratic_voteSS=group_frac(IL_state_map,adv_18,(adv_18+arv_18)))

hist(m_illinois_plans_2020,Population_deviation) + hist(m_illinois_plans_2020,Compactness)#+
  #plot_layout(guides="collect")+
  #plot_annototion(title="Simulated plan charachteristics")
```
```{r post-run analysis-illinois 02}
redist.plot.scatter(m_illinois_plans_2020, Population_deviation,Compactness)+
  labs(title="Population deviation and compactness by plan")
```
```{r post-run analysis-illinois 03}
plot(m_illinois_plans_2020,Democratic_vote,size=0.5,color_thresh=0.5)+
  scale_color_manual(values=c("black","tomato2","dodgerblue"))+
  labs(title="Democratic vote share by district")
```


Texas saved diagnostic information, just for tuning purposes.

With 2000 samples:

Plan diversity 80% range: 0.21 to 1.10
✖ WARNING: Low plan diversity

R-hat values for summary statistics:
pop_overlap     pop_dev        comp     pct_min     pct_dem 
    ❌1.074       1.043     ❌1.151     ❌1.101     ❌1.243 
✖ WARNING: SMC runs have not converged.

Sampling diagnostics for SMC run 1 of 2 (2,000 samples)

Sampling diagnostics for SMC run 2 of 2 (2,000 samples)

For 2 runs of 4000:
Plan diversity 80% range: 0.1 to 1.1
✖ WARNING: Low plan diversity

R-hat values for summary statistics:
pop_overlap     pop_dev        comp     pct_min     pct_dem 
    ❌1.224     ❌1.703     ❌2.181     ❌2.125     ❌2.291 
✖ WARNING: SMC runs have not converged.

Sampling diagnostics for SMC run 1 of 2 (4,000 samples)

Sampling diagnostics for SMC run 2 of 2 (4,000 samples)



#Ilinois info saved
SMC: 4,000 sampled plans of 17 districts on 10,084 units
`adapt_k_thresh`=0.99 • `seq_alpha`=0.5
`est_label_mult`=1 • `pop_temper`=0

Plan diversity 80% range: 1.0 to 1.1

R-hat values for summary statistics:
pop_overlap     pop_dev        comp     pct_min     pct_dem 
      1.046       1.038     ❌1.093       1.005       1.004 
✖ WARNING: SMC runs have not converged.

Sampling diagnostics for SMC run 1 of 2 (2,000 samples)

Sampling diagnostics for SMC run 2 of 2 (2,000 samples)

•  Watch out for low effective samples, very low acceptance rates (less than 1%), large std.
devs. of the log weights (more than 3 or so), and low numbers of unique plans. R-hat values
for summary statistics should be between 1 and 1.05.
• SMC convergence: Increase the number of samples. If you are experiencing low plan
diversity or bottlenecks as well, address those issues first.

SMC: 8,000 sampled plans of 17 districts on 10,084 units
`adapt_k_thresh`=0.99 • `seq_alpha`=0.5
`est_label_mult`=1 • `pop_temper`=0

#

Plan diversity 80% range: 1.0 to 1.1

R-hat values for summary statistics:
pop_overlap     pop_dev        comp     pct_min     pct_dem 
      1.021       1.013       1.032     ❌1.055       1.044 
✖ WARNING: SMC runs have not converged.

Sampling diagnostics for SMC run 1 of 2 (4,000 samples)

Sampling diagnostics for SMC run 2 of 2 (4,000 samples)

•  Watch out for low effective samples, very low acceptance rates (less than 1%), large std.
devs. of the log weights (more than 3 or so), and low numbers of unique plans. R-hat values
for summary statistics should be between 1 and 1.05.
• SMC convergence: Increase the number of samples. If you are experiencing low plan
diversity or bottlenecks as well, address those issues first.

#

SMC: 16,000 sampled plans of 17 districts on 10,084 units
`adapt_k_thresh`=0.99 • `seq_alpha`=0.5
`est_label_mult`=1 • `pop_temper`=0

Plan diversity 80% range: 0.99 to 1.11

R-hat values for summary statistics:
pop_overlap     pop_dev        comp     pct_min     pct_dem 
    ❌1.065       1.028       1.009       1.014       1.010 
✖ WARNING: SMC runs have not converged.

Sampling diagnostics for SMC run 1 of 2 (8,000 samples)

Sampling diagnostics for SMC run 2 of 2 (8,000 samples)

I do not want to throw out my code for texas yet

```{r texas code}
#TX_state_map<-alarm_50state_map('TX')
# print("Checking Texas:")
# dim_to_check<-dim(TX_state_map)
# print("The dimensions of the Texas input data is: ")
# print(dim_to_check)
# 
# # count total missing values 
# print("Count of total missing values in Texas Data:  ")
# sum(is.na(TX_state))
# 
# # find location of missing values
# print("Position of missing values ")
# which(is.na(TX_state))

```
```{r map01-texas}
# #Texas
# areas= as.numeric(units::set_units(sf::st_area(TX_state_map$geometry),mi^2))
# plot(TX_state_map, adj=T)
# 
# TX_map_plot<-TX_state_map |>
#   ggplot() +
#   geom_sf(aes(fill = ndv / (nrv+ndv)))+
#   scale_fill_party_c() +
#   theme_map()
# TX_map_plot


```

maps will enable us to analyze our proposed maps against randomly generated plans. 
```{r redist maps-texas}
#Texas
# texas_plans<-redist_smc(TX_state_map,1000,compactness = 1,runs = 2 )
# print(texas_plans)

```

```{r sample plots-texas}
#redist.plot.plans(texas_plans, draws=1:6, shp=TX_state_map)
```


```{r 2020 comparison - texas}
# #Matching district numbers
# #texas_2010_map<-match_numbers(texas_plans,TX_state_map$cd_2010)
# texas_plans_2020<-match_numbers(texas_plans,TX_state_map$cd_2020)
# #print(texas_plans)
# #Adding summary statistics for both Census
# #county_perims2010<- prep_perims(texas_2010_map, TX_state_map$adj)
# county_perims2020<- prep_perims(TX_state_map, TX_state_map$adj)
# 
# #we will be analyze the population deviation, the perimeter-based compactness measure, and the fraction of minority voters and two-party Democratic vote share by district.
# 
# texas_plans_2020 = texas_plans_2020 %>%
#   mutate(pop_dev = abs(total_pop / get_target(TX_state_map)-1),
#          comp=comp_polsby(pl(), TX_state_map,perim_df = county_perims2020),
#          pct_min = group_frac(TX_state_map, vap-vap_white,vap),
#          pct_dem= group_frac(TX_state_map,adv_18,adv_18+arv_18))
# 
# print(texas_plans_2020)

```


```{r Diagnostics-texas}
#summary(texas_plans_2020)

```



